\documentclass{article}
\usepackage{amsmath} % For mathematical equations
\usepackage{geometry} % For adjusting margins
\usepackage{amssymb} % For sets (\mathbb{R}, etc. if needed, and \emptyset)
\usepackage{amsthm} % For definition environment if needed

% Adjust page margins
\geometry{a4paper, margin=1in}

% Improve paragraph spacing
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0em}

\title{Mathematical Optimization Model for Weekly Task Scheduling}
\author{Formulation Documentation}
\date{\today}

\begin{document}

\maketitle

\section{Overview}

This document presents the mathematical formulation of an optimization model designed to generate a weekly task schedule. The model assigns start times to tasks within a defined time horizon, respecting various constraints, and optimizes a weighted combination of maximizing leisure time and minimizing task-related stress.

\subsection*{Time Horizon}
The planning horizon covers $D = 7$ days. Each day consists of $S_{day} = 56$ discrete time slots of 15 minutes each, typically representing the period from 8:00 AM to 10:00 PM. The total number of slots in the horizon is $S_{total} = D \times S_{day} = 392$. Slots are indexed globally from $s = 0$ to $S_{total}-1$.

\section{Parameters and Inputs}

The model utilizes the following input parameters:

\begin{itemize}
    \item \textbf{Set of Tasks ($T$):} The collection of tasks to be scheduled, indexed by $i$.
    \item \textbf{Task Attributes ($\forall i \in T$):}
    \begin{itemize}
        \item $p_i$: Numerical priority of task $i$.
        \item $d_i$: Numerical difficulty of task $i$.
        \item $dur_i$: Duration of task $i$, measured in the number of time slots.
        \item $dl_i$: Deadline for task $i$, represented as the global slot index by which the task must be fully completed.
        \item $AllowedSlots_i$: A subset of $\{0, ..., S_{total}-1\}$ indicating the permissible start slots for task $i$ based on its time preference (e.g., "morning", "afternoon").
    \end{itemize}
    \item \textbf{Set of Committed Slots ($C$):} A subset of $\{0, ..., S_{total}-1\}$ representing time slots that are pre-allocated and unavailable for scheduling tasks (e.g., meetings, appointments).
    \item \textbf{Objective Weights:}
    \begin{itemize}
        \item $\alpha \ge 0$: Weight coefficient emphasizing the maximization of leisure time.
        \item $\beta \ge 0$: Weight coefficient emphasizing the minimization of the total stress score.
    \end{itemize}
    \item \textbf{Daily Limit ($Limit_{daily}$, Optional):} An integer representing the maximum number of slots that can be occupied by tasks on any single day $d \in \{0, ..., D-1\}$.
\end{itemize}

\section{Decision Variables}

The model determines the values of the following variables:

\begin{enumerate}
    \item \textbf{$X_{i,s}$ (Binary):} Indicates if task $i$ starts at slot $s$.
    \[ X_{i,s} = \begin{cases} 1 & \text{if task } i \in T \text{ starts at slot } s \in \{0, ..., S_{total}-1\} \\ 0 & \text{otherwise} \end{cases} \]

    \item \textbf{$Y_{s}$ (Binary):} Indicates if slot $s$ is occupied by any task.
    \[ Y_{s} = \begin{cases} 1 & \text{if slot } s \in \{0, ..., S_{total}-1\} \text{ is occupied by a task} \\ 0 & \text{otherwise} \end{cases} \]
    This is an auxiliary variable linked to $X_{i,s}$ via constraints.

    \item \textbf{$L_{s}$ (Continuous):} Represents the leisure time (in minutes) available in slot $s$.
    \[ L_{s} \in [0, 15] \quad \forall s \in \{0, ..., S_{total}-1\} \]
\end{enumerate}

\section{Objective Function}

The objective is to maximize a weighted sum reflecting the trade-off between leisure and stress:

\[
\text{Maximize} \quad Z = \alpha \sum_{s=0}^{S_{total}-1} L_s - \beta \sum_{i \in T} \sum_{s=0}^{S_{total}-1} (p_i \times d_i) X_{i,s}
\]

\textbf{Explanation:}
\begin{itemize}
    \item The first term, $\alpha \sum L_s$, promotes maximizing the total leisure time across all slots. The weight $\alpha$ scales the importance of leisure relative to stress reduction.
    \item The second term, $\beta \sum (p_i \times d_i) X_{i,s}$, aims to minimize the total "stress" incurred. Stress for each task $i$ is defined as $p_i \times d_i$ and is counted if the task starts ($X_{i,s}=1$). The weight $\beta$ scales the penalty associated with scheduling high-priority or difficult tasks. This formulation associates stress with the initiation of a task, not its entire duration.
\end{itemize}

\section{Constraints}

The following constraints define the feasible schedules:

\subsection{Task Assignment}
Ensures that each task is assigned exactly one starting slot within the horizon.

\[
\sum_{s=0}^{S_{total}-1} X_{i,s} = 1 \quad \forall i \in T
\]

\textbf{Explanation:} For every task $i$, the sum of its binary start variables $X_{i,s}$ over all possible slots $s$ must equal 1. This forces the model to select exactly one slot $s$ where $X_{i,s}=1$ (the start slot) for each task $i$.

\subsection{Deadlines and Horizon}
Ensures that tasks are completed by their deadline and fit entirely within the scheduling horizon.

\[
X_{i,s} = 0 \quad \forall i \in T, \forall s \text{ such that } (s + dur_i - 1 > dl_i) \lor (s + dur_i > S_{total})
\]

\textbf{Explanation:}
\begin{itemize}
    \item \textbf{Deadline Check ($s + dur_i - 1 > dl_i$):} A task $i$ starting at slot $s$ occupies slots from $s$ to $s + dur_i - 1$. The term $s + dur_i - 1$ represents the index of the last slot occupied by the task. This constraint forces the start variable $X_{i,s}$ to be 0 if the last occupied slot exceeds the task's specified deadline slot $dl_i$. This prevents tasks from starting if they cannot finish on time.
    \item \textbf{Horizon Check ($s + dur_i > S_{total}$):} The term $s + dur_i$ represents the index of the slot immediately following the task's completion. If this index is greater than the total number of slots $S_{total}$, it means the task would extend beyond the planning horizon. This constraint forces $X_{i,s}$ to 0 in such cases, ensuring tasks are fully contained within the schedule.
\end{itemize}

\subsection{No Overlap}
Prevents two or more tasks from being active in the same time slot.

\[
\sum_{i \in T} \sum_{s = \max(0, t - dur_i + 1)}^{t} X_{i,s} \le 1 \quad \forall t \in \{0, ..., S_{total}-1\}
\]

\textbf{Explanation:} For any given time slot $t$, this constraint considers all tasks $i$. The inner sum iterates through potential start slots $s$ for task $i$. The range $\max(0, t - dur_i + 1)$ to $t$ identifies precisely those start slots $s$ such that task $i$, if started at $s$, would be active during slot $t$. By summing the $X_{i,s}$ variables for all such task/start-slot combinations, we count how many tasks are active in slot $t$. Constraining this sum to be less than or equal to 1 ensures that at most one task can occupy any single slot $t$.

\subsection{Preferences}
Restricts the starting time of tasks to their allowed time windows (e.g., morning, afternoon).

\[
X_{i,s} = 0 \quad \forall i \in T, \forall s \notin AllowedSlots_i
\]

\textbf{Explanation:} For each task $i$, $AllowedSlots_i$ is the set of permissible start slots based on user preference. This constraint directly enforces the preference by setting the start variable $X_{i,s}$ to 0 for any slot $s$ that is not within the task's allowed set. This prevents the task from starting outside its desired time window(s).

\subsection{Commitments}
Prevents any part of a scheduled task from coinciding with predefined fixed commitments.

\[
X_{i,s} = 0 \quad \forall i \in T, \forall s \text{ such that } \{s, s+1, ..., s + dur_i - 1\} \cap C \neq \emptyset
\]

\textbf{Explanation:} For each task $i$ and potential start slot $s$, the set $\{s, s+1, ..., s + dur_i - 1\}$ represents all slots the task would occupy if it started at $s$. This constraint checks if this set has any overlap (non-empty intersection, $\cap C \neq \emptyset$) with the set of committed slots $C$. If an overlap exists, it means starting task $i$ at $s$ would conflict with a commitment. Therefore, the constraint forces $X_{i,s}$ to 0, preventing the task from starting at that time.

\subsection{Leisure Calculation and Occupation Link (Y)}
Links the auxiliary occupation variable $Y_s$ to the task start variables $X_{i,s}$ and defines the leisure time $L_s$ based on occupation and commitments.

\begin{align}
Y_s &= \sum_{i \in T} \sum_{start = \max(0, s - dur_i + 1)}^{s} X_{i,start} \quad &&\forall s \in \{0, ..., S_{total}-1\} \label{eq:link_y} \\
L_s &= 0 \quad &&\forall s \in C \label{eq:leisure_commit} \\
L_s &\le 15 \times (1 - Y_s) \quad &&\forall s \notin C \label{eq:leisure_task} \\
L_s &\ge 0 \quad &&\forall s \label{eq:leisure_nonneg}
\end{align}

\textbf{Explanation:}
\begin{itemize}
    \item \textbf{Equation (\ref{eq:link_y}):} This constraint defines the occupation variable $Y_s$. It equates $Y_s$ to the same sum used in the "No Overlap" constraint. Since the No Overlap constraint ensures that this sum can only be 0 or 1, this equality correctly sets $Y_s = 1$ if slot $s$ is occupied by any task, and $Y_s = 0$ otherwise. This makes $Y_s$ a reliable binary indicator of task occupation for slot $s$.
    \item \textbf{Equation (\ref{eq:leisure_commit}):} If slot $s$ belongs to the set of commitments $C$, no leisure time is possible, so $L_s$ is forced to 0.
    \item \textbf{Equation (\ref{eq:leisure_task}):} If slot $s$ is not committed ($s \notin C$), this constraint limits the leisure time. If the slot is occupied by a task ($Y_s = 1$), then $1 - Y_s = 0$, forcing $L_s \le 0$. Since $L_s$ must also be non-negative (Eq. \ref{eq:leisure_nonneg}), $L_s$ becomes exactly 0. If the slot is not occupied by a task ($Y_s = 0$), then $1 - Y_s = 1$, allowing $L_s \le 15$. The objective function, seeking to maximize $\sum L_s$, will push $L_s$ to its upper bound of 15 in this unoccupied, non-committed case.
    \item \textbf{Equation (\ref{eq:leisure_nonneg}):} Explicitly states (or is implied by variable definition) that leisure time cannot be negative.
\end{itemize}

\subsection{Daily Limits (Optional)}
Enforces a maximum total duration of tasks scheduled within any single day.

Let $Slots_{day, d} = \{s \mid d \times S_{day} \le s < (d+1) \times S_{day}\}$ be the set of slots for day $d$.
\[
\sum_{s \in Slots_{day, d}} Y_s \le Limit_{daily} \quad \forall d \in \{0, ..., D-1\}
\]

\textbf{Explanation:} This constraint operates on a per-day basis ($d=0$ to $D-1$). For each day, it sums the occupation variables $Y_s$ over all slots $s$ belonging to that day ($Slots_{day, d}$). Since $Y_s=1$ if a slot is occupied by a task and 0 otherwise, this sum represents the total number of task-occupied slots on day $d$. The constraint ensures this total does not exceed the specified $Limit_{daily}$.

\section{Output}

If the optimization problem is feasible and a solution is found (potentially optimal or feasible within a time limit), the model output provides the values of the decision variables. This typically translates to:
\begin{itemize}
    \item The optimal objective function value $Z$.
    \item A schedule detailing the assigned start slot $s$ (where $X_{i,s}=1$) for each task $i$.
    \item The calculated total leisure time $\sum L_s$.
    \item The calculated total stress score $\sum (p_i \times d_i) X_{i,s}$.
\end{itemize}

\end{document}
